---
title: "Getting started with prakriti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with prakriti}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

## What is prakriti?

**prakriti** (Sanskrit for *nature*) provides 30 hand-tuned color palettes
inspired by India's natural landscapes. Each palette is designed for a specific
use case - sequential, diverging, or qualitative - and integrates directly with
`ggplot2`.

```{r load}
library(prakriti)
library(ggplot2)
```

## Discovering palettes

```{r names}
prakriti_names()
```

Get structured metadata with `prakriti_info()`:

```{r info}
prakriti_info()
```

## Viewing palettes

Display a single palette:

```{r display-one, fig.height=1.5}
display_prakriti("valley_of_flowers")
```

Display all 30 palettes at once:

```{r display-all, fig.height=12}
display_prakriti()
```

## Using palettes directly

`prakriti_palette()` returns a character vector of hex codes:

```{r palette-basic}
prakriti_palette("himalaya")
prakriti_palette("thar", n = 3)
```

Reverse with `direction = -1`:

```{r palette-reverse}
prakriti_palette("chinar", direction = -1)
```

Interpolate to any number of colors with `type = "continuous"`:

```{r palette-continuous}
prakriti_palette("nilgiri", n = 12, type = "continuous")
```

## ggplot2 integration

### Qualitative palettes (categorical data)

```{r scatter, fig.height=4}
ggplot(iris, aes(Sepal.Length, Petal.Length,
                 color = Species, shape = Species)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_prakriti("valley_of_flowers") +
  labs(title = "Iris dataset - valley_of_flowers palette") +
  theme_minimal()
```

### Sequential palettes (continuous data)

```{r heatmap}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_prakriti("himalaya") +
  coord_cartesian(expand = FALSE) +
  labs(title = "Old Faithful - himalaya palette") +
  theme_minimal()
```

### Diverging palettes (data with a meaningful midpoint)

```{r diverging}
cor_mat <- cor(mtcars[, 1:7])
cor_df  <- as.data.frame(as.table(cor_mat))
names(cor_df) <- c("var1", "var2", "rho")

ggplot(cor_df, aes(var1, var2, fill = rho)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_prakriti("mehrangarh", limits = c(-1, 1)) +
  coord_equal() +
  labs(title = "Correlation matrix - mehrangarh palette") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Overriding defaults

By default, qualitative palettes produce discrete scales and sequential/diverging
produce continuous scales. Override with `discrete`:

```{r override}
# Force a sequential palette into discrete mode
ggplot(mtcars, aes(factor(cyl), mpg, fill = factor(cyl))) +
  geom_boxplot() +
  scale_fill_prakriti("thar", discrete = TRUE) +
  labs(title = "Forced discrete: thar palette") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Full tutorial

For 24 detailed plot examples covering every palette type - stacked area,
donut charts, calendar heatmaps, correlation tiles, dark-mode density fields,
and more - see `tutorial.R` in the package root.
